Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow

### 2024-04-09 manuscript/Manuscript.qmd

---
title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
This recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to religably estimate an indicator value for each EA.
This problem does not exist for complete wall-to-wall data like remotely sensed imagery where each EA will get its own unbiased indicator value.

There are at least 2 ways to achieve this same complete areal coverage, that is relatively unbiased, from originally point-based indicator values:

a.  Predict (project) values from a statistical model that accounts for the effect of environmental variation on the variables
b.  Simple projection of some best estimate, for example a central tendency from a stratified selection of area representative data

@@

This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature [@norwegian_environmental_agency_protocol].
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and are therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
For example, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
In this study we focused on open mire ecosystems in Norway where the thematic bias is less severe.
The survey maps the following mire types which we use in our analyses:

@@

Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
These ranges were chosen as they produced a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each range across Norway.

The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
We plotted the HIA map over an area in Norway that we are familiar with and confirmed a sensible separation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e. the ecosystem accounting areas).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

>>>>

---
title: Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to religably estimate an indicator value for each EA.
This problem does not exist for complete wall-to-wall data like remotely sensed imagery where each EA will get its own unbiased indicator value.

There are at least 3 ways to achieve this same complete areal coverage (i.e. spatialisation) that is relatively unbiased, from originally point-based indicator values:

a.  Predict (project) values from a statistical model that accounts for the effect of environmental variation on the variables
b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
c.  Take a central tendency from stratified sections of the sample polulation and projection them to the EAs, adhering to the corresponding strata.

This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].

@@

This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature [@norwegian_environmental_agency_protocol].
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and are therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
For example, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
In this study we focused on open mire ecosystems in Norway where the thematic bias is less severe (the spatial bias being presumably equal for all ecosystems).
The survey maps the following mire types which we use in our analyses:

@@

Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
These ranges were chosen as they produced a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each range across Norway.
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.

The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
We refer to this step as the first of two *aggregation* steps (Fig. S1).
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
This step is referred to as *spreading* the data (Fig. S1).
For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Fig. S1).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.
